sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2022-11-14 doc/paper-jss/paper-jss.Rmd
Edti

---
documentclass: jss
author:
  - name: Sean C. Anderson
    orcid: 0000-0001-9563-1937
    affiliation: 'Pacific Biological Station'
    address: |
      | Fisheries and Oceans Canada, Nanaimo, BC, Canada
    email: \email{sean.anderson@dfo-mpo.gc.ca}
  - name: Eric J. Ward
    orcid: 0000-0002-4359-0296
    affiliation: 'Northwest Fisheries Science Center \AND'
    address: |
      | National Oceanic and Atmospheric Administration, National Marine Fisheries Service, Seattle, WA, USA
  - name: Philina A. English
    orcid: 0000-0003-2992-6782
    affiliation: 'Pacific Biological Station'
    address: |
      | Fisheries and Oceans Canada, Nanaimo, BC, Canada
  - name: Lewis A. K. Barnett
    orcid: 0000-0002-9381-8375
    affiliation: 'Alaska Fisheries Science Center'
    address: |
      | National Marine Fisheries Service, National Oceanic and Atmospheric Administration, Seattle, WA, USA
bibliography: refs.bib
title:
  formatted: "\\pkg{sdmTMB}: An \\proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # If you use tex in the formatted title, also supply version without
  plain:     "sdmTMB: An R Package for Fast, Flexible, and Accessible Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # For running headers, if needed
  short:     "\\pkg{sdmTMB}: geostatistical SPDE-based GLMMs with \\pkg{TMB}"
abstract: >
  Geostatistical data are common across scientific fields and available in increasingly large datasets. However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to implement for many users. Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach. SPDE matrices are constructed with \proglang{R}-\pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}) or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}. We describe the model and explore case studies that illustrate \pkg{sdmTMB}'s flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (asymmetry in spatial correlations). Additional functionality includes hurdle models, break-point effects, and out-of-sample cross validation. Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that sdmTMB can be 4–16 times faster than \proglang{R}-\pkg{INLA}. We hope \pkg{sdmTMB}'s accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
keywords:
  # at least one keyword must be supplied
  formatted: [Gaussian Markov random fields (GMRF), generalized linear mixed effects model (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
  plain:     [Gaussian Markov random fields (GMRF), generalized linear mixed effects model (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{bm}
  \usepackage{lscape}
  \usepackage{tablefootnote}
  \usepackage{threeparttable}
  \usepackage{booktabs}
  \usepackage{pifont}
  \usepackage{newunicodechar}
  \newunicodechar{✓}{\ding{51}}
  \newunicodechar{✗}{\ding{55}}
  \DeclareGraphicsExtensions{.pdf,.png,.jpg}
output: rticles::jss_article
---

@@

Data are often collected in space or in space repeatedly over time.
While such data are a rich source of information about mechanistic processes [@legendre1989a; @rossi1992; @tilman1997spatial], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart due to measured and unmeasured variables [@cressie1993; @diggle2007; @cressie2011].
While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
Accounting for this residual correlation is important because doing so allows for valid statistical inference [@legendre1989a; @dormann2007], can improve predictions  [e.g., @shelton2014], and can be of interest itself by, for example, identifying locations with shared responses [e.g., @thorson2019d; @barnett2021].

@@

Such models quickly become computationally limiting due to the need to invert large matrices to keep track of covariation among data [e.g., @rue2005gmrf; @latimer2009].
Many solutions have been proposed, such as predictive processes [@banerjee2008; @latimer2009], the stochastic partial differential equation (SPDE) approximation to GRFs [@lindgren2011], and nearest-neighbour Gaussian processes [@datta2016; @finley2021].
These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian [@gelfand2017] or maximum likelihood methods.
This can greatly improve computational efficiency [e.g., @heaton2019].
The SPDE approach is a solution popularized via the \pkg{INLA} \proglang{R} package [@rue2009; @lindgren2011; @lindgren2015] and an implementation in \pkg{TMB} [Template Model Builder, @kristensen2016] that partially relies on \pkg{INLA} to create input matrices [e.g., @osgood-zimmerman2021].
Details are beyond the scope of this paper and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this 'trick' enables one to efficiently fit approximations to GRFs to large spatial datasets [@lindgren2011].

Systems for specifying statistical models that can include the SPDE, such as \pkg{INLA} and \pkg{TMB}, are flexible and powerful but are challenging to use for many applied researchers.
For example, \pkg{TMB} requires the user to program in a \proglang{C++} template and it can be slow to experiment with multiple models when writing bespoke model code.
Packages such as \pkg{lme4} [@bates2015] and \pkg{glmmTMB} [@brooks2017] let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---but do not have built-in SPDE functionality.
Packages such as \pkg{VAST} [@thorson2019] and \pkg{inlabru} [@bachl2019] are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used \pkg{lme4} or \pkg{glmmTMB}, or lack some functionality.
We provide a more detailed comparison of related software packages in Table \ref{tab:functionality} and the Discussion.

Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which implements geostatistical spatial and spatiotemporal GLMMs using \pkg{TMB} for model fitting and \pkg{INLA} to set up SPDE matrices.
Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} [@wood2017a], for a specific class of spatial and spatiotemporal models.
One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \pkg{TMB}: sdmTMB).
This paper describes the basic functionality of this \proglang{R} package and its underlying statistical model, illustrates its use through two case studies, and concludes with a comparison to related software.

@@

The GLMMs underpinning \pkg{sdmTMB} models are spatially explicit---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
In contrast, semi- and non-parametric approaches do not estimate spatial covariance functions (e.g., randomForest [@liaw2002], MaxEnt [@phillips2006], and most smooths in \pkg{mgcv} [@wood2017a]).
The random fields in \pkg{sdmTMB} are structured as MVN constrained by a Matérn covariance function [@matern1960].
The Matérn can accommodate a range of shapes and can be both isotropic (covariance decays the same in all directions) or anisotropic (covariance in the latitudinal direction may differ from the longitudinal direction) [@haskard2007].
The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to $\sim 0.13$ [@lindgren2015]---can be shared or estimated separately (`share_range` argument).

By default, if spatiotemporal fields are included, they are assumed IID; however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \ref{fig:diagram}, Appendix \ref{app:model}).
Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in \pkg{mgcv} [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes; for example, hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()` (Appendix \ref{app:model}).
The smoothers in \pkg{sdmTMB} can mimic most `mgcv::s()` and `mgcv::t2()` smoothers including bivarate smoothers (`s(x, y)`), smoothers varying by continuous or categorical variables (`s(x1, by = x2)`), cyclical smoothers `s(x, bs = "cc")`, and smoothers with specified basis dimensions `s(x, k = 4)` [@wood2017a].

The \pkg{sdmTMB} model is fit by maximum marginal likelihood.
Internally, a \pkg{TMB} [@kristensen2016] model template is used to calculate the marginal log likelihood and its gradient, and the negative log likelihood is minimized via the non-linear optimization routine `stats::nlminb()` in \proglang{R} [@gay1990; @r2021].
Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].
After rapid model exploration with maximum likelihood, one can optionally pass an \pkg{sdmTMB} model to the \proglang{R} package \pkg{tmbstan} [@monnahan2018] to estimate the joint posterior distribution for Bayesian inference.

@@

<!-- An \pkg{sdmTMB} model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates. -->
Usually it makes sense to convert the spatial coordinates to an equidistant projection such as the Universal Transverse Mercator (UTM) to ensure that distance remains constant throughout the study region [e.g., using `sf::st_transform()`, @pebesma2018, or `add_utm_columns()`].
Here we use `add_utm_columns()` to add UTM coordinates with km units (so our estimated spatial range parameter is not too big or small).
By default, the function will guess our UTM zone and create new columns `X` and `Y`.

@@

Here, `cutoff` defines the minimum allowed distance between mesh vertices in the units of `X` and `Y`.
`make_mesh()` is a wrapper function for `INLA::inla.mesh.create()`.
Alternatively, we could have created any mesh via the \pkg{INLA} package and supplied it to the `mesh` argument in `make_mesh()`.
We can inspect our mesh object with the associated plotting method (`plot(mesh)`; see Fig. \ref{fig:pcod-fig}A).

@@

The output indicates our model was fit by maximum (marginal) likelihood (`ML`).
We also see the formula, mesh, fitted data, and family.
Next we see any estimated main effects including the linear component of the smoother (`sdepth`), the standard deviation on the smoother weights (`sds(depth)`), the Tweedie dispersion and power parameters, the Matérn range distance, the marginal spatial field standard deviation, and the negative log likelihood at convergence.
The smoother summary follows the format used in the \pkg{brms} package [@brms].
As noted in the summary method, we can extract a tidy data frame similar to using the \pkg{broom} [@broom] or \pkg{broom.mixed} [@broommixed] packages:

@@

We can make predictions with the `predict()` method (`?predict.sdmTMB`) and optionally use the `newdata` argument to predict on a new data frame with any locations and values for the predictor columns.
Here, we will predict on a 2x2 km grid (`qcs_grid`) that covers the entire region of interest so we can visualize the predictions spatially (or calculate a standardized population index with a spatiotemporal model).
The grid contains spatial covariate columns and all predictors used in the model set at values for which we want to predict.

@@

The output of `predict()` is a data frame containing overall estimates in link space (`est`), estimates from the non-random-field components (intercept and depth) (`est_non_rf`), and estimates from the random field components combined (`est_rf`), and the individual random field components (here `omega_s`---the spatial field).

@@

We show a basic plot of the estimated spatial random field, predictions across a depth gradient, and predictions in space in Fig. \ref{fig:pcod-fig}B--D.

@@

We could inspect the anisotropy with `plot_anisotropy()`, which plots an ellipse showing the spatial range with the spatial coordinates centered on zero.
We could generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Fig. \ref{fig:pcod-fig}E, Appendix \ref{app:pcod}).

@@

Validation and selection of state-space models is challenging, particularly when using the Laplace approximation [@thygesen2017a].
We provide several approaches to assist this process:
(1) The Akaike Information Criterion [AIC, @akaike1974] can be calculated with `AIC()`, although AIC has well-documented biases with mixed-effects models [@liang2008].
(2) Alternatively, k-fold cross validation with `sdmTMB_cv()` can be used with user-specified or randomly chosen folds for model selection [e.g., via expected log predictive density [ELPD], @vehtari2017] or to evaluate goodness of fit according to user-calculated criteria (e.g., mean squared error, area under the curve).
(3) An \pkg{sdmTMB} model can be passed to the \pkg{tmbstan} package [@monnahan2018] to sample from the joint posterior with Stan [@carpenter2017], evaluate the accuracy of the Laplace approximation, or perform posterior predictive checks (see `?extract_mcmc`).
(4) The `residuals()` method by default returns randomized quantile [@dunn1996] or probability integral transform (PIT) [@smith1985] residuals.
<!-- These residuals, however, are calculated as $\hat{y} - y$ where $\hat{y}$ is the sum of the fixed effect estimates and the random effect values that maximize the likelihood conditional on the estimated fixed effects.  -->
For state-space models, these residuals have known statistical issues with the Laplace approximation [@thygesen2017a] but are quick to calculate.
A version that uses Markov chain Monte Carlo (MCMC) to avoid this issue is recommended but slower (`?residuals.sdmTMB`).
Simulation-based residuals [@dharma] (`dharma_residuals()`) are also possible.
(5) The `simulate.sdmTMB()` method can simulate from fitted models and the `sdmTMB_simulate()` function can simulate entirely new data to which models can be fit to ensure identifiability, evaluate bias and precision in parameter estimation, or evaluate the consequences of model misspecification.

@@

There are many \proglang{R} packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2019].
\pkg{sdmTMB}, \pkg{VAST}, \pkg{INLA}/\pkg{inlabru}, and \pkg{spaMM} [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GRF models; we also include \pkg{spBayes} in our software comparison as it is a prominent package that can fit related predictive-process models without the SPDE (Table \ref{tab:functionality}).
\pkg{sdmTMB}, \pkg{VAST}, and \pkg{mgcv} can estimate anisotropic covariance whereas \pkg{INLA}/\pkg{inlabru} and \pkg{spBayes} are currently limited to isotropic covariance.
\pkg{sdmTMB} and \pkg{mgcv} focus on univariate response data, whereas \pkg{VAST}, \pkg{INLA}/\pkg{inlabru}, \pkg{spaMM}, and \pkg{spBayes} extend to multivariate responses with various limitations.
To our knowledge, \pkg{VAST} is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016] and spatial empirical orthogonal function (EOF) regression [@thorson2020].
Of these packages, only \pkg{sdmTMB} and \pkg{inlabru} can currently fit threshold (e.g., hockey-stick) covariate relationships.
\pkg{spaMM} is limited to a spatial random field and \pkg{spBayes} implements spatiotemporal fields, but only as a random walk.
There is considerable variability in the available observation likelihoods across packages (Table \ref{tab:functionality}).
We provide comparisons of the syntax and the reproducibility of results from models fit using \pkg{INLA} and \pkg{inlabru} in Appendix \ref{app:inla}.
<!-- and \pkg{VAST} (Appendix 5). -->

@@

How does one choose among the related packages mentioned in this paper to fit SPDE-based geostatistical GLMMs?
Assuming a given package can fit the model of interest (Table \ref{tab:functionality}), we suggest the primary differences are the user interface (Appendix \ref{app:inla}) and speed.
We think users familiar with `stats::glm()`, \pkg{lme4}, or \pkg{glmmTMB} will find \pkg{sdmTMB} most approachable.
Users familiar with \pkg{INLA} will find \pkg{inlabru} approachable.
Users familiar with \pkg{mgcv} can adapt \pkg{mgcv} to fit similar models with custom code [@miller2019] and \pkg{INLA}/\pkg{inlabru} and \pkg{mgcv} are also general purpose modelling packages.
\pkg{VAST} is the sole option for fitting some multivariate models;
alternatively, because \pkg{VAST} focuses on multivariate delta models and fisheries applications, users fitting "simple" univariate spatial/spatiotemporal GLMMs in non-fisheries contexts may find \pkg{sdmTMB} more straightforward to use.
Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour \pkg{sdmTMB} or \pkg{VAST} (although such quantities can be post hoc derived with other packages).
<!-- Speed-wise, the TMB-based packages \pkg{sdmTMB} and \pkg{VAST} were fastest up to at least 1000 mesh nodes, which in practice can cover fairly complex spatial processes on large datasets [e.g., @anderson2019b; @maureaud2021; @english2022]. -->

>>>>

---
documentclass: jss
author:
  - name: Sean C. Anderson
    orcid: 0000-0001-9563-1937
    affiliation: 'Pacific Biological Station'
    address: |
      | Fisheries and Oceans Canada, Nanaimo, BC, Canada
    email: \email{sean.anderson@dfo-mpo.gc.ca}
  - name: Eric J. Ward
    orcid: 0000-0002-4359-0296
    affiliation: 'Northwest Fisheries Science Center \AND'
    address: |
      | National Oceanic and Atmospheric Administration, National Marine Fisheries Service, Seattle, WA, USA
  - name: Philina A. English
    orcid: 0000-0003-2992-6782
    affiliation: 'Pacific Biological Station'
    address: |
      | Fisheries and Oceans Canada, Nanaimo, BC, Canada
  - name: Lewis A. K. Barnett
    orcid: 0000-0002-9381-8375
    affiliation: 'Alaska Fisheries Science Center'
    address: |
      | National Marine Fisheries Service, National Oceanic and Atmospheric Administration, Seattle, WA, USA
bibliography: refs.bib
title:
  formatted: "\\pkg{sdmTMB}: An \\proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # If you use tex in the formatted title, also supply version without
  plain:     "sdmTMB: An R Package for Fast, Flexible, and Accessible Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # For running headers, if needed
  short:     "\\pkg{sdmTMB}: geostatistical SPDE-based GLMMs with \\pkg{TMB}"
abstract: >
  Geostatistical data are common across scientific fields and available in increasingly large datasets. However, appropriate models to analyse these data, such as generalised linear mixed effects models (GLMMs) with Gaussian random fields, are computationally intensive and challenging to implement for many users. Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach. SPDE matrices are constructed with \proglang{R}-\pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}) or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}. We describe the model and explore case studies that illustrate \pkg{sdmTMB}'s flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (asymmetry in spatial correlations). Additional functionality includes hurdle models, break-point effects, and out-of-sample cross validation. Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that sdmTMB can be 4–16 times faster than \proglang{R}-\pkg{INLA}. We hope \pkg{sdmTMB}'s accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
keywords:
  # at least one keyword must be supplied
  formatted: [Gaussian Markov random fields (GMRF), generalized linear mixed effects model (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
  plain:     [Gaussian Markov random fields (GMRF), generalized linear mixed effects model (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{bm}
  \usepackage{lscape}
  \usepackage{tablefootnote}
  \usepackage{threeparttable}
  \usepackage{booktabs}
  \usepackage{pifont}
  \usepackage{newunicodechar}
  \newunicodechar{✓}{\ding{51}}
  \newunicodechar{✗}{\ding{55}}
  \DeclareGraphicsExtensions{.pdf,.png,.jpg}
  \widowpenalty10000
  \clubpenalty10000
output: rticles::jss_article
---

@@

Data are often collected in space or in space repeatedly over time.
While such data are a rich source of information [e.g., @legendre1989a; @rossi1992; @tilman1997spatial], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart due to measured and unmeasured variables [@cressie1993; @diggle2007; @cressie2011].
While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
Accounting for this residual correlation is important because doing so allows for valid statistical inference [@legendre1989a; @dormann2007], can improve predictions  [e.g., @shelton2014], and can be of interest itself by, for example, identifying locations with shared responses [e.g., @thorson2019d; @barnett2021].

@@

Such models quickly become computationally limiting due to the need to invert large matrices to keep track of covariation among data [e.g., @rue2005gmrf; @latimer2009].
Many solutions have been proposed, such as predictive processes [@banerjee2008; @latimer2009], the stochastic partial differential equation (SPDE) approximation to GRFs [@lindgren2011], and nearest-neighbour Gaussian processes [@datta2016; @finley2021].
These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian [@gelfand2017] or maximum likelihood methods.
This can greatly improve computational efficiency [e.g., @heaton2019].
The SPDE approach is a solution popularized via the \proglang{R}-\pkg{INLA} \proglang{R} package [@rue2009; @lindgren2011; @lindgren2015] and an implementation in \pkg{TMB} [Template Model Builder, @kristensen2016] that partially relies on \proglang{R}\-pkg{INLA} to create input matrices [e.g., @thorson2019; @osgood-zimmerman2021].
Details are beyond the scope of this paper and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this 'trick' enables one to efficiently fit approximations to GRFs to large spatial datasets [@lindgren2011].

Systems for specifying statistical models that can include the SPDE, such as \proglang{R}-\pkg{INLA} and \pkg{TMB}, are flexible and powerful but are challenging to use for many applied researchers.
For example, \pkg{TMB} requires the user to program in a \proglang{C++} template and it can be slow to experiment with multiple models when writing bespoke model code.
Packages such as \pkg{lme4} [@bates2015] and \pkg{glmmTMB} [@brooks2017] let users quickly iterate and explore statistical models---focusing on evaluating fit and comparing models---but do not have built-in SPDE functionality.
Packages such as \pkg{VAST} [@thorson2019] and \pkg{inlabru} [@bachl2019] are powerful user interfaces to fit spatial models that use the SPDE, but they either lack a modular interface familiar to those who have used \pkg{lme4} or \pkg{glmmTMB}, or lack some functionality.
We provide a more detailed comparison of related software packages in Table \ref{tab:functionality} and the Discussion.

Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which implements geostatistical spatial and spatiotemporal GLMMs using \pkg{TMB} for model fitting and \proglang{R}-\pkg{INLA} to set up SPDE matrices.
Our aim is not to replace the above-mentioned statistical packages, but to provide a fast, flexible, and user-friendly interface that is familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} [@wood2017a], for a specific class of spatial and spatiotemporal models.
One common application in the field of ecology is species distribution models (SDMs), hence the package name (i.e., SDMs with \pkg{TMB}: sdmTMB).
This paper describes the basic functionality of this \proglang{R} package and its underlying statistical model, illustrates its use through two case studies, and concludes with a comparison to related software.

@@

The GLMMs underpinning \pkg{sdmTMB} models are spatially explicit---they estimate interpretable parameters of a spatial covariance function: parameters defining the magnitude of spatial variation and the rate of correlation decay with distance.
In contrast, semi- and non-parametric approaches do not estimate spatial covariance functions (e.g., randomForest [@liaw2002], MaxEnt [@phillips2006], and most smooths in \pkg{mgcv} [@wood2017a]).
The random fields in \pkg{sdmTMB} are structured as MVN constrained by a Matérn covariance function [@matern1960].
The Matérn can accommodate a range of shapes and can be both isotropic (covariance decays the same in all directions) or anisotropic (covariance is directionally dependent) [@haskard2007].
The Matérn standard deviations are estimated separately for the various fields and the range---the distance at which spatial correlation decays to $\sim 0.13$ [@lindgren2015]---can be shared or estimated separately (`share_range` argument).

By default, if spatiotemporal fields are included, they are assumed IID; however, additional options allow them to be modelled as a random walk or first-order autoregressive, AR(1), process (Fig.\ \ref{fig:diagram}, Appendix \ref{app:model}).
Turning off both spatial and spatiotemporal effects allows comparison with a standard non-spatial GLM or GLMM.
We include additional flexibility in specifying the linear fixed effect matrix: covariates can be modelled as penalized smooth functions (generalized additive models, GAMs) using the same `s()` syntax as in \pkg{mgcv} [@wood2017a] (thereby allowing automatic selection of smoother 'wiggliness'), or can be modelled with threshold shapes; for example, hockey-stick models, `breakpt()`, [@barrowman2000] or logistic functions, `logistic()` (Appendix \ref{app:model}).
The smoothers in \pkg{sdmTMB} can mimic most `mgcv::s()` and `mgcv::t2()` smoothers including bivarate smoothers (`s(x, y)`), smoothers varying by continuous or categorical variables (`s(x1, by = x2)`), cyclical smoothers (`s(x, bs = "cc")`), and smoothers with specified basis dimensions (`s(x, k = 4)`) [@wood2017a].

The \pkg{sdmTMB} model is fit by maximum marginal likelihood.
Internally, a \pkg{TMB} [@kristensen2016] model template is used to calculate the marginal log likelihood and its gradient, and the negative log likelihood is minimized via the non-linear optimization routine `stats::nlminb()` in \proglang{R} [@gay1990; @r2021].
Random effects are estimated at values that maximize the log likelihood conditional on the estimated fixed effects and are integrated over via the Laplace approximation [@kristensen2016].
After rapid model exploration with maximum likelihood, one can optionally pass an \pkg{sdmTMB} model to the \proglang{R} package \pkg{rstan} [@carpenter2017; @rstan] via \pkg{tmbstan} [@monnahan2018] to estimate the joint posterior distribution for Bayesian inference.

@@

<!-- An \pkg{sdmTMB} model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates. -->
Usually it makes sense to convert the spatial coordinates to an equidistant projection such as the Universal Transverse Mercator (UTM) to ensure that distance remains constant throughout the study region [e.g., using `sf::st_transform()`, @pebesma2018].
Here we use the helper function `add_utm_columns()` to add UTM coordinates with km units (so our estimated spatial range parameter is not too big or small).
By default, the function will guess our UTM zone and create new columns `X` and `Y`.

@@

Here, `cutoff` defines the minimum allowed distance between mesh vertices in the units of `X` and `Y`.
`make_mesh()` is a wrapper function for `INLA::inla.mesh.create()`.
Alternatively, we could have created any mesh via the \proglang{R}-\pkg{INLA} package and supplied it to the `mesh` argument in `make_mesh()`.
We can inspect our mesh object with the associated plotting method (`plot(mesh)`; see Fig. \ref{fig:pcod-fig}A).

@@

The output indicates our model was fit by maximum (marginal) likelihood (`ML`).
We also see the formula, mesh, fitted data, and family.
Next we see any estimated main effects including the linear component of the smoother (`sdepth`), the standard deviation on the smoother weights (`sds(depth)`), the Tweedie dispersion and power parameters, the Matérn range distance, the marginal spatial field standard deviation, and the negative log likelihood at convergence.
The smoother summary follows the format used in the \pkg{brms} package [@burkner2017].
As noted in the summary method, we can extract a tidy data frame similar to using the \pkg{broom} [@broom] or \pkg{broom.mixed} [@broommixed] packages:

@@

We can make predictions with the `predict()` method (`?predict.sdmTMB`) and optionally use the `newdata` argument to predict on a new data frame with any locations and values for the predictor columns.
Here, we will predict on a 2 $\times$ 2 km grid (`qcs_grid`) that covers the entire region of interest so we can visualize the predictions spatially (or calculate a standardized population index with a spatiotemporal model).
The grid contains spatial covariate columns and all predictors used in the model set at values for which we want to predict.

@@

The output of `predict()` is a data frame containing overall estimates in link space (`est`), estimates from the non-random-field components (intercept and depth) (`est_non_rf`), and estimates from the random field components combined (`est_rf`) and the individual random field components (here `omega_s`---the spatial field).

@@

We show a basic plot of the estimated spatial random field, predictions across a depth gradient, and predictions in space in Fig.\ \ref{fig:pcod-fig}B--D.

@@

We could inspect the anisotropy with `plot_anisotropy()`, which plots an ellipse showing the spatial range with the spatial coordinates centered on zero.
We could generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Fig.\ \ref{fig:pcod-fig}E, Appendix \ref{app:pcod}).

@@

Validation and selection of state-space models is challenging, particularly when using the Laplace approximation [@thygesen2017a].
We provide several approaches to assist this process:
(1) The Akaike Information Criterion [AIC, @akaike1974] can be calculated with `AIC()`, although AIC has well-documented biases with mixed-effects models [@liang2008].
(2) Alternatively, k-fold cross validation with `sdmTMB_cv()` can be used with user-specified or randomly chosen folds for model selection [e.g., via expected log predictive density [ELPD], @vehtari2017] or to evaluate goodness of fit according to user-calculated criteria (e.g., mean squared error, area under the curve).
(3) An \pkg{sdmTMB} model can be passed to the \pkg{tmbstan} package [@monnahan2018] to sample from the joint posterior with Stan [@carpenter2017], evaluate the accuracy of the Laplace approximation, or perform posterior predictive checks (see `?extract_mcmc`).
(4) The `residuals()` method by default returns randomized quantile [@dunn1996] or probability integral transform (PIT) [@smith1985] residuals.
<!-- These residuals, however, are calculated as $\hat{y} - y$ where $\hat{y}$ is the sum of the fixed effect estimates and the random effect values that maximize the likelihood conditional on the estimated fixed effects.  -->
For state-space models, these residuals have known statistical issues with the Laplace approximation [@thygesen2017a] but are quick to calculate.
A version that uses Markov chain Monte Carlo (MCMC) to avoid this issue is recommended but slower (`?residuals.sdmTMB`).
Simulation-based residuals [@dharma] (`dharma_residuals()`) are also possible.
(5) The `simulate.sdmTMB()` method can simulate from fitted models and the `sdmTMB_simulate()` function facilitates simulating data without starting from a fitted model. 
Models can be fit to these simulated data to ensure identifiability, evaluate bias and precision in parameter estimation, or evaluate the consequences of model misspecification.

@@

There are many \proglang{R} packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2019].
\pkg{sdmTMB}, \pkg{VAST}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, and \pkg{spaMM} [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GRF models; we also include \pkg{spBayes} in our software comparison as it is a prominent package that can fit related predictive-process models without the SPDE (Table \ref{tab:functionality}).
\pkg{sdmTMB}, \pkg{VAST}, and \pkg{mgcv} can estimate anisotropic covariance whereas \proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{spBayes} are currently limited to isotropic covariance.
\pkg{sdmTMB} and \pkg{mgcv} focus on univariate response data, whereas \pkg{VAST}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, \pkg{spaMM}, and \pkg{spBayes} extend to multivariate responses with various limitations.
To our knowledge, \pkg{VAST} is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016] and spatial empirical orthogonal function (EOF) regression [@thorson2020].
Of these packages, only \pkg{sdmTMB} and \pkg{inlabru} can currently fit threshold (e.g., hockey-stick) covariate relationships.
\pkg{spaMM} is limited to a spatial random field and \pkg{spBayes} implements spatiotemporal fields, but only as a random walk.
There is considerable variability in the available observation likelihoods across packages (Table \ref{tab:functionality}).
We provide comparisons of the syntax and the reproducibility of results from models fit using \proglang{R}-\pkg{INLA} and \pkg{inlabru} in Appendix \ref{app:inla}.
<!-- and \pkg{VAST} (Appendix 5). -->

@@

How does one choose among the related packages mentioned in this paper to fit SPDE-based geostatistical GLMMs?
Assuming a given package can fit the model of interest (Table \ref{tab:functionality}), we suggest the primary differences are the user interface (Appendix \ref{app:inla}) and speed.
We think users familiar with `stats::glm()`, \pkg{lme4}, \pkg{mgcv}, or \pkg{glmmTMB} will find \pkg{sdmTMB} most approachable.
Users familiar with \proglang{R}-\pkg{INLA} will find \pkg{inlabru} approachable.
Users familiar with \pkg{mgcv} can adapt \pkg{mgcv} to fit similar models with custom code [@miller2019] and \proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{mgcv} are also general purpose modelling packages.
\pkg{VAST} is the sole option for fitting some multivariate models;
alternatively, because \pkg{VAST} focuses on multivariate delta models and fisheries applications, users fitting "simple" univariate spatial/spatiotemporal GLMMs in non-fisheries contexts may find \pkg{sdmTMB} more straightforward to use.
Users looking for calculation, with uncertainty, of derived variables such as area-weighted population indexes, may favour \pkg{sdmTMB} or \pkg{VAST} (although such quantities can be post hoc derived with other packages).
<!-- Speed-wise, the TMB-based packages \pkg{sdmTMB} and \pkg{VAST} were fastest up to at least 1000 mesh nodes, which in practice can cover fairly complex spatial processes on large datasets [e.g., @anderson2019b; @maureaud2021; @english2022]. -->
